---
title: "01 - Load Data & Baseline QC"
author: "Adrienne Yanez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

Before applying any QC filtering or clustering, obtain a baseline read on
the raw Xenium output: how many cells were called, and how many transcripts
and genes were detected per cell. Xenium's targeted panel is far sparser
than scRNA-seq (median ~49 transcripts and ~46 genes/cell here), so these
baseline distributions are what determine the three QC filtering thresholds
compared in `03_qc_comparison.Rmd`.

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_knit$set(root.dir = here::here())
library(Seurat)
library(ggplot2)
library(patchwork)
library(here)
library(arrow)

# Data directory is not tracked in this repo.
# Set XENIUM_DATA_DIR in a local .Renviron to override.
# See data/download_instructions.md to download the dataset.
data_dir <- Sys.getenv(
  "XENIUM_DATA_DIR",
  unset = file.path(path.expand("~"), "Documents", "one-sample-analysis",
                    "Xenium_Prime_Breast_Cancer_FFPE_outs")
)

stopifnot(
  "Data directory not found — see data/download_instructions.md" = file.exists(data_dir)
)

xenium.obj <- LoadXenium(data_dir, fov = "fov")
```

```{r qc, message=FALSE, warning=FALSE}
# Violin plots of per-cell transcript/gene counts, unfiltered. These
# distributions are the basis for the lenient/standard/strict thresholds

# "standard" (nCount > 25, nFeature > 20) sits just above
# the median so it removes the sparse tail without discarding typical cells
p <- VlnPlot(xenium.obj, features = c("nCount_Xenium", "nFeature_Xenium"), ncol = 2, pt.size = 0)
ggsave(here("figures", "qc_violin_unfiltered.png"), plot = p, width = 10, height = 5, dpi = 300)

# baseline stats
qc_summary <- data.frame(
  metric = c("Total cells (unfiltered)", "Total genes", "Median counts per cell", "Median genes per cell"),
  value  = c(
    ncol(xenium.obj),
    nrow(xenium.obj),
    median(xenium.obj$nCount_Xenium),
    median(xenium.obj$nFeature_Xenium)
  )
)
write.csv(qc_summary, here("figures", "qc_summary_unfiltered.csv"), row.names = FALSE)

print(qc_summary)
```
