---
title: "02 - Clustering"
author: "Adrienne Yanez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

This notebook establishes the **standard filtering** baseline clustering result
(nCount > 25, nFeature > 20, 454,652 cells retained). It is the primary
reference point that `03_qc_comparison.Rmd` measures lenient and strict
filtering against.

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_knit$set(root.dir = here::here())

library(Seurat)
library(ggplot2)
library(patchwork)
library(here)
library(arrow)
# library(BPCells)  # needed only if re-running sketch pipeline from scratch


data_dir <- Sys.getenv(
  "XENIUM_DATA_DIR",
  unset = file.path(path.expand("~"), "Documents", "one-sample-analysis",
                    "Xenium_Prime_Breast_Cancer_FFPE_outs")
)

stopifnot(
  "Data directory not found - see data/download_instructions.md" = file.exists(data_dir)
)

# CATALYST color palette - shared across all notebooks for consistent cluster colors
catalyst_cols <- c(
  "#7BAFDE", "#B2DF8A", "#BEAED4", "#8DD3C7", "#DC050C",
  "#FF7F00", "#882E72", "#33A02C", "#E7298A", "#E6AB02",
  "#1965B0", "#55A1B1", "#7570B3",
  "#FB8072", "#E78AC3", "#B17BA6", "#FDB462", "#A6761D", "#666666"
)

# Factor cluster column in numerical order and return named palette
fix_clusters <- function(obj, col = "cluster_full") {
  vals <- as.character(obj[[col, drop = TRUE]])
  lvls <- as.character(sort(as.integer(unique(vals))))
  obj[[col]] <- factor(vals, levels = lvls)
  pal <- setNames(catalyst_cols[seq_along(lvls)], lvls)
  list(obj = obj, pal = pal, levels = lvls)
}
```

---

## 1. Load Data

```{r load-data, message=FALSE}
# Standard filtering (nCount > 25, nFeature > 20) is the baseline clustering
# result for this project. It retains 454,652 cells and serves
# as the reference point for the QC threshold comparison in 03_qc_comparison.Rmd.
# Pipeline was run in a prior session. See sketch chunk below for details.
xenium.obj <- readRDS(here("data", "xenium_standard.rds"))

# Re-level cluster_full (stored as character after ProjectData)
res        <- fix_clusters(xenium.obj, col = "cluster_full")
xenium.obj <- res$obj
pal_full   <- res$pal

# Rebuild sketch palette from seurat_clusters
sketch_levels <- as.character(sort(as.integer(unique(
  as.character(xenium.obj$seurat_clusters)
))))
xenium.obj$seurat_clusters <- factor(xenium.obj$seurat_clusters,
                                      levels = sketch_levels)
pal_sketch <- setNames(catalyst_cols[seq_along(sketch_levels)], sketch_levels)

cat("Cells:", ncol(xenium.obj), "\n")
cat("Clusters:", length(res$levels), "\n")
```

---

## 2. Sketch: Subsample for Memory-Efficient Analysis

The full dataset (~699k cells) exceeds available RAM for ScaleData.
Subsample 50,000 representative cells using leverage-score sketching,
run the full pipeline on the sketch, then project labels back to all cells.
50,000 is Seurat 5's optimal size for sketch-based workflows because it is
large enough to capture rare populations via leverage scoring, and still fits PCA/clustering in memory on a laptop.

```{r sketch, message=FALSE, eval=FALSE}
# -----------------------------------------------------------------------------
# NOTE: This chunk is set to eval=FALSE - the pre-computed object is loaded
# above. To reproduce from scratch, set eval=TRUE and run in a clean R session
# immediately after LoadXenium() with no intervening commands.
#
# KNOWN ISSUE: SketchData() with LeverageScore attempts a sparse to dense matrix
# cast (~26.6 GiB) that exceeds 32GB RAM on this 699k cell dataset. The saved
# object (data/xenium_standard.rds) was generated using random subsampling
# (set.seed(42), n=50,000) as a memory-efficient alternative. Results are
# equivalent for clustering purposes; leverage scoring would better preserve
# rare populations but requires >32GB RAM or a BPCells matrix.
# -----------------------------------------------------------------------------

# To reproduce from scratch:
xenium.obj <- LoadXenium(data_dir, fov = "fov")

set.seed(42)
cells_keep  <- sample(Cells(xenium.obj), size = 50000)
xenium.obj  <- subset(xenium.obj, cells = cells_keep)
cat("Subsampled cells:", ncol(xenium.obj), "\n")

DefaultAssay(xenium.obj) <- "Xenium"
xenium.obj <- NormalizeData(xenium.obj)
xenium.obj <- FindVariableFeatures(xenium.obj)
xenium.obj <- ScaleData(xenium.obj, features = VariableFeatures(xenium.obj))
xenium.obj <- RunPCA(xenium.obj, npcs = 50)
xenium.obj <- FindNeighbors(xenium.obj, dims = 1:25)
xenium.obj <- FindClusters(xenium.obj, resolution = 0.3)
xenium.obj <- RunUMAP(xenium.obj, dims = 1:25, return.model = TRUE)

xenium.obj <- ProjectData(
  object             = xenium.obj,
  assay              = "Xenium",
  full.reduction     = "full.pca",
  sketched.assay     = "sketch",
  sketched.reduction = "pca",
  umap.model         = "umap",
  dims               = 1:25,
  refdata            = list(cluster_full = "seurat_clusters")
)

saveRDS(xenium.obj, here("data", "xenium_standard.rds"))
```

---

## 3. Preprocessing

```{r preprocess, message=FALSE, eval=FALSE}
# eval=FALSE - preprocessing was run as part of the pipeline that generated
# data/xenium_standard.rds. Shown here for documentation only.
xenium.obj <- NormalizeData(xenium.obj)
xenium.obj <- FindVariableFeatures(xenium.obj)
xenium.obj <- ScaleData(xenium.obj, features = VariableFeatures(xenium.obj))
```

---

## 4. Dimensionality Reduction

```{r pca, message=FALSE, eval=FALSE}
# eval=FALSE because PCA was run as part of the pipeline that generated
# data/xenium_standard.rds. Shown here for documentation only.
xenium.obj <- RunPCA(xenium.obj, npcs = 50)
```

```{r elbow-plot, fig.width=7, fig.height=4}
# PCA already exists in the loaded object. Regenerate elbow plot for display
ElbowPlot(xenium.obj, ndims = 50) +
  ggtitle("Elbow Plot - PCA") +
  theme_minimal()
```

```{r save-02-elbow, include=FALSE}
ggsave(here("figures", "02_elbow_plot.png"),
       ElbowPlot(xenium.obj, ndims = 50) +
         ggtitle("Elbow Plot - PCA") + theme_minimal(),
       width = 7, height = 4, dpi = 300)
```

> **Note:** `dims` in downstream steps set to 1:25 based on elbow inflection.

---

## 5. Clustering

```{r cluster, message=FALSE, eval=FALSE}
# eval=FALSE - clustering was run as part of the pipeline that generated
# data/xenium_standard.rds. Shown here for documentation only.
xenium.obj <- FindNeighbors(xenium.obj, dims = 1:25)
xenium.obj <- FindClusters(xenium.obj, resolution = 0.3)
xenium.obj <- RunUMAP(xenium.obj, dims = 1:25, return.model = TRUE)
```

```{r cluster-summary}
# Summarise clusters from the loaded object
sketch_levels <- as.character(sort(as.integer(levels(xenium.obj$seurat_clusters))))
pal_sketch <- setNames(catalyst_cols[seq_along(sketch_levels)], sketch_levels)
cat("Clusters:", length(sketch_levels), "\n")
table(xenium.obj$seurat_clusters)
```

---

## 6. UMAP

```{r umap-plot, fig.width=7, fig.height=6}
DimPlot(xenium.obj, reduction = "umap", label = TRUE,
        label.size = 4, cols = pal_sketch) +
  ggtitle("UMAP - Clusters (sketch, n=50k)") +
  theme_minimal()
```

```{r save-02-umap, include=FALSE}
ggsave(here("figures", "02_umap_clusters.png"),
       DimPlot(xenium.obj, reduction = "umap", label = TRUE,
               label.size = 4, cols = pal_sketch) +
         ggtitle("UMAP - Clusters (sketch, n=50k)") + theme_minimal(),
       width = 7, height = 6, dpi = 300)
```

---

## 7. Project Back to All Cells

```{r project, message=FALSE, eval=FALSE}
# eval=FALSE because cluster_full was projected when xenium_standard.rds was 
# generated. Re-projection happens automatically on load via fix_clusters()
xenium.obj <- ProjectData(
  object             = xenium.obj,
  assay              = "Xenium",
  full.reduction     = "full.pca",
  sketched.assay     = "sketch",
  sketched.reduction = "pca",
  umap.model         = "umap",
  dims               = 1:25,
  refdata            = list(cluster_full = "seurat_clusters")
)

cat("Projected cluster labels to all", ncol(xenium.obj), "cells\n")

# Factor cluster_full numerically and build palette for all-cell plots
res        <- fix_clusters(xenium.obj, col = "cluster_full")
xenium.obj <- res$obj
pal_full   <- res$pal
cat("Cluster levels (full):", res$levels, "\n")
```

---

## 8. Spatial Plot

```{r spatial-plot, fig.width=10, fig.height=8, message=FALSE, warning=FALSE}
DefaultAssay(xenium.obj) <- "Xenium"

ImageDimPlot(xenium.obj,
             group.by = "cluster_full",
             cols     = pal_full,
             size     = 0.5) +
  ggtitle("Spatial Clusters - Standard Filtering Baseline")
```

```{r save-02-spatial, include=FALSE}
ggsave(here("figures", "02_spatial_clusters.png"),
       ImageDimPlot(xenium.obj,
                    group.by = "cluster_full",
                    cols     = pal_full,
                    size     = 0.5) +
         ggtitle("Spatial Clusters - Unfiltered Dataset"),
       width = 10, height = 8, dpi = 300)
```

---

## 9. Side-by-Side: UMAP + Spatial

```{r combined-plot, fig.width=16, fig.height=7, message=FALSE, warning=FALSE}
DefaultAssay(xenium.obj) <- "sketch"
p1 <- DimPlot(xenium.obj, reduction = "umap", label = TRUE,
              label.size = 4, cols = pal_sketch) +
  ggtitle("UMAP") + theme_minimal() + NoLegend()

DefaultAssay(xenium.obj) <- "Xenium"
p2 <- ImageDimPlot(xenium.obj,
                   group.by = "cluster_full",
                   cols     = pal_full,
                   size     = 0.5) +
  ggtitle("Spatial") + NoLegend()

p1 + p2
```

```{r save-02-combined, include=FALSE}
ggsave(here("figures", "02_umap_spatial_combined.png"),
       p1 + p2,
       width = 16, height = 7, dpi = 300)
```

---

## 10. Figures Generated

| File | Description |
|------|-------------|
| `02_elbow_plot.png` | PCA elbow plot |
| `02_umap_clusters.png` | UMAP colored by cluster (sketch) |
| `02_spatial_clusters.png` | Spatial clusters, full dataset |
| `02_umap_spatial_combined.png` | UMAP + spatial side by side |

---

## 11. Session Info

```{r session-info}
sessionInfo()
```
