---
title: "03 - QC Threshold Comparison"
author: "Adrienne Yanez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

This notebook is the core comparison to address how QC filtering
stringency changes which cell populations are recovered from Xenium data.
Three thresholds (lenient/standard/strict) are each run through
the identical clustering pipeline from `02_clustering.Rmd` so cluster
counts and spatial distributions are directly comparable, isolating the
effect of the QC cutoff itself from any other  variation.

```{r setup, message=FALSE, warning=FALSE}

knitr::opts_knit$set(root.dir = here::here())

library(Seurat)
library(ggplot2)
library(patchwork)
library(here)
library(arrow)

data_dir <- Sys.getenv(
  "XENIUM_DATA_DIR",
  unset = file.path(path.expand("~"), "Documents", "one-sample-analysis",
                    "Xenium_Prime_Breast_Cancer_FFPE_outs")
)

catalyst_cols <- c(
  "#7BAFDE", "#B2DF8A", "#BEAED4", "#8DD3C7", "#DC050C",
  "#FF7F00", "#882E72", "#33A02C", "#E7298A", "#E6AB02",
  "#1965B0", "#55A1B1", "#7570B3",
  "#FB8072", "#E78AC3", "#B17BA6", "#FDB462", "#A6761D", "#666666"
)

fix_clusters <- function(obj, col = "cluster_full") {
  vals <- as.character(obj[[col, drop = TRUE]])
  lvls <- as.character(sort(as.integer(unique(vals))))
  obj[[col]] <- factor(vals, levels = lvls)
  pal <- setNames(catalyst_cols[seq_along(lvls)], lvls)
  list(obj = obj, pal = pal, levels = lvls)
}

```


```{r load-objects, message=FALSE}
# Load objects sequentially and extract only what's needed for plotting.
# Full objects are ~1.5GB each - loading all three simultaneously hits 32GB RAM.
# Extract cluster metadata only, then reload full objects one at a time for plots.

obj_lenient  <- readRDS(here("data", "xenium_lenient.rds"))
meta_lenient <- data.frame(
  cluster_full = as.character(obj_lenient$cluster_full),
  nCount       = obj_lenient$nCount_Xenium,
  row.names    = Cells(obj_lenient)
)
n_lenient <- ncol(obj_lenient)
rm(obj_lenient); gc()

obj_standard <- readRDS(here("data", "xenium_standard.rds"))
meta_standard <- data.frame(
  cluster_full = as.character(obj_standard$cluster_full),
  nCount       = obj_standard$nCount_Xenium,
  row.names    = Cells(obj_standard)
)
n_standard <- ncol(obj_standard)
rm(obj_standard); gc()

obj_strict <- readRDS(here("data", "xenium_strict.rds"))
meta_strict <- data.frame(
  cluster_full = as.character(obj_strict$cluster_full),
  nCount       = obj_strict$nCount_Xenium,
  row.names    = Cells(obj_strict)
)
n_strict <- ncol(obj_strict)
rm(obj_strict); gc()

# shared cluster levels across all conditions for color consistency
shared_levels <- as.character(sort(as.integer(unique(c(
  meta_lenient$cluster_full,
  meta_standard$cluster_full,
  meta_strict$cluster_full
)))))
shared_pal <- setNames(catalyst_cols[seq_along(shared_levels)], shared_levels)

cat("Lenient:", n_lenient, "cells,", 
    length(unique(meta_lenient$cluster_full)), "clusters\n")
cat("Standard:", n_standard, "cells,", 
    length(unique(meta_standard$cluster_full)), "clusters\n")
cat("Strict:", n_strict, "cells,", 
    length(unique(meta_strict$cluster_full)), "clusters\n")
cat("Shared cluster IDs:", shared_levels, "\n")
```


---

## Overview

**Question:** How much do QC filtering thresholds affect the clusters and
recovered from Xenium data?

Three filtering strategies are compared, plus the unfiltered baseline from
notebook 02:

| Strategy    | nCount_Xenium | nFeature_Xenium | Rationale |
|-------------|--------------|-----------------|-----------|
| Unfiltered  | none         | none            | Baseline, notebook 02 |
| Lenient     | > 10         | > 10            | Minimal filtering |
| Standard    | > 25         | > 20            | Near-median and balanced tradeoffs|
| Strict      | > 100        | > 50            | High quality cells only |

Baseline: **699,110 cells**, median counts = 49, median genes = 46.

Thresholds are set relative to baseline. 
- "lenient" only removes near-empty cells (counts/features in the single
digits, likely background or segmentation noise)
- "standard" is just below the dataset median, so it drops the sparsest cells
- "strict" keeps only cells well above median, trading cell
yield for panel coverage

---

## 1. Load Raw Data

```{r load data, message=FALSE, eval=FALSE}
xenium.obj <- LoadXenium(data_dir, fov = "fov")
cat("Raw cells:", ncol(xenium.obj), "\n")
```

---

## 2. Define Pipeline Function

```{r pipeline function, eval=FALSE}
# Identical to the notebook 02 pipeline (50k random subsample, set.seed(42),
# resolution = 0.3, dims = 1:25 from the elbow plot). NOTE: eval=FALSE,
# saved objects are loaded directly above. SketchData() with LeverageScore
# was the original approach but exceeds 32GB RAM on this dataset.

run_pipeline <- function(obj, label, dims = 1:25) {
  message("\nRunning pipeline: ", label)
  message("Cells: ", ncol(obj))

 set.seed(42)
  cells_keep <- sample(Cells(obj), size = min(50000, ncol(obj)))
  obj <- subset(obj, cells = cells_keep)
  message("Subsampled to: ", ncol(obj), " cells")

  obj <- NormalizeData(obj)
  obj <- FindVariableFeatures(obj)
  obj <- ScaleData(obj)
  obj <- RunPCA(obj, npcs = 50)
  obj <- FindNeighbors(obj, dims = dims)
  obj <- FindClusters(obj, resolution = 0.3)
  obj <- RunUMAP(obj, dims = dims, return.model = TRUE)

  obj <- ProjectData(
    object             = obj,
    assay              = "Xenium",
    full.reduction     = "full.pca",
    sketched.reduction = "pca",
    umap.model         = "umap",
    dims               = dims,
    refdata            = list(cluster_full = "seurat_clusters")
  )

  vals <- as.character(obj$cluster_full)
  lvls <- as.character(sort(as.integer(unique(vals))))
  obj$cluster_full <- factor(vals, levels = lvls)

  message("Clusters recovered: ", length(lvls))
  return(obj)
}
```

---

## 3. Filter and run each  condition

### 3.1 Lenient (nCount > 10, nFeature > 10)

```{r lenient, message=TRUE, warning=FALSE, eval=FALSE}
obj_lenient <- subset(xenium.obj,
                      subset = nCount_Xenium > 10 & nFeature_Xenium > 10)
cat("Lenient - cells retained:", ncol(obj_lenient),
    sprintf("(%.1f%% of raw)\n", 100 * ncol(obj_lenient) / ncol(xenium.obj)))

obj_lenient <- run_pipeline(obj_lenient, label = "Lenient")
```

### 3.2 Standard (nCount > 25, nFeature > 20)

```{r standard, message=TRUE, warning=FALSE, eval=FALSE}
obj_standard <- subset(xenium.obj,
                       subset = nCount_Xenium > 25 & nFeature_Xenium > 20)
cat("Standard - cells retained:", ncol(obj_standard),
    sprintf("(%.1f%% of raw)\n", 100 * ncol(obj_standard) / ncol(xenium.obj)))

obj_standard <- run_pipeline(obj_standard, label = "Standard")
```

### 3.3 Strict (nCount > 100, nFeature > 50)

```{r strict, message=TRUE, warning=FALSE, eval=FALSE}
obj_strict <- subset(xenium.obj,
                     subset = nCount_Xenium > 100 & nFeature_Xenium > 50)
cat("Strict - cells retained:", ncol(obj_strict),
    sprintf("(%.1f%% of raw)\n", 100 * ncol(obj_strict) / ncol(xenium.obj)))

obj_strict <- run_pipeline(obj_strict, label = "Strict")
```


---

## 4. Summary Table

```{r summary-table}
qc_comparison <- data.frame(
  condition      = c("Unfiltered", "Lenient", "Standard", "Strict"),
  min_counts     = c(NA, 10, 25, 100),
  min_features   = c(NA, 10, 20, 50),
  cells_retained = c(699110, n_lenient, n_standard, n_strict),
  pct_of_raw     = round(100 * c(699110, n_lenient, n_standard, n_strict) / 699110, 1),
  n_clusters     = c(
    NA,
    length(unique(meta_lenient$cluster_full)),
    length(unique(meta_standard$cluster_full)),
    length(unique(meta_strict$cluster_full))
  )
)

print(qc_comparison)
write.csv(qc_comparison, here("figures", "03_qc_comparison_summary.csv"),
          row.names = FALSE)
```

---

## 5. UMAPs

```{r umap-comparison, fig.width=18, fig.height=6, message=FALSE, warning=FALSE}
# Load, plot remove each object sequentially to stay within 32GB RAM

obj_lenient <- readRDS(here("data", "xenium_lenient.rds"))
obj_lenient$cluster_full <- factor(as.character(obj_lenient$cluster_full),
                                    levels = shared_levels)
p_lenient <- DimPlot(obj_lenient, reduction = "umap", label = TRUE,
                     cols = shared_pal, label.size = 3, pt.size = 0.3) +
  ggtitle(sprintf("Lenient\n(n=%s, %d clusters)",
                  format(n_lenient, big.mark = ","),
                  sum(table(obj_lenient$cluster_full) > 0))) +
  theme_minimal() + NoLegend()
rm(obj_lenient); gc()

obj_standard <- readRDS(here("data", "xenium_standard.rds"))
obj_standard$cluster_full <- factor(as.character(obj_standard$cluster_full),
                                     levels = shared_levels)
p_standard <- DimPlot(obj_standard, reduction = "umap", label = TRUE,
                      cols = shared_pal, label.size = 3, pt.size = 0.3) +
  ggtitle(sprintf("Standard\n(n=%s, %d clusters)",
                  format(n_standard, big.mark = ","),
                  sum(table(obj_standard$cluster_full) > 0))) +
  theme_minimal() + NoLegend()
rm(obj_standard); gc()

obj_strict <- readRDS(here("data", "xenium_strict.rds"))
obj_strict$cluster_full <- factor(as.character(obj_strict$cluster_full),
                                   levels = shared_levels)
p_strict <- DimPlot(obj_strict, reduction = "umap", label = TRUE,
                    cols = shared_pal, label.size = 3, pt.size = 0.3) +
  ggtitle(sprintf("Strict\n(n=%s, %d clusters)",
                  format(n_strict, big.mark = ","),
                  sum(table(obj_strict$cluster_full) > 0))) +
  theme_minimal() + NoLegend()
rm(obj_strict); gc()

p_lenient + p_standard + p_strict
```

```{r save-03-umap-comparison, include=FALSE}
ggsave(here("figures", "03_qc_comparison_umaps.png"),
       p_lenient + p_standard + p_strict,
       width = 18, height = 6, dpi = 300)
```

---

## 6. Spatial Plots

```{r spatial-comparison, fig.width=18, fig.height=7, message=FALSE, warning=FALSE}
# Load, plot, save, remove each object sequentially to stay within 32GB RAM.
# ImageDimPlot is memory-intensive - do not hold all three objects simultaneously.

obj_lenient <- readRDS(here("data", "xenium_lenient.rds"))
obj_lenient$cluster_full <- factor(as.character(obj_lenient$cluster_full),
                                    levels = shared_levels)
DefaultAssay(obj_lenient) <- "Xenium"
sp_lenient <- ImageDimPlot(obj_lenient, group.by = "cluster_full",
                           cols = shared_pal, size = 0.4) +
  ggtitle("Lenient Spatial") + NoLegend()
ggsave(here("figures", "03_spatial_lenient.png"),
       sp_lenient, width = 8, height = 7, dpi = 300)
rm(obj_lenient); gc()

obj_standard <- readRDS(here("data", "xenium_standard.rds"))
obj_standard$cluster_full <- factor(as.character(obj_standard$cluster_full),
                                     levels = shared_levels)
DefaultAssay(obj_standard) <- "Xenium"
sp_standard <- ImageDimPlot(obj_standard, group.by = "cluster_full",
                            cols = shared_pal, size = 0.4) +
  ggtitle("Standard Spatial") + NoLegend()
ggsave(here("figures", "03_spatial_standard.png"),
       sp_standard, width = 8, height = 7, dpi = 300)
rm(obj_standard); gc()

obj_strict <- readRDS(here("data", "xenium_strict.rds"))
obj_strict$cluster_full <- factor(as.character(obj_strict$cluster_full),
                                   levels = shared_levels)
DefaultAssay(obj_strict) <- "Xenium"
sp_strict <- ImageDimPlot(obj_strict, group.by = "cluster_full",
                          cols = shared_pal, size = 0.4) +
  ggtitle("Strict Spatial") + NoLegend()
ggsave(here("figures", "03_spatial_strict.png"),
       sp_strict, width = 8, height = 7, dpi = 300)
rm(obj_strict); gc()

# Combine and display - ggplot objects are lightweight, safe to combine
sp_lenient + sp_standard + sp_strict
```

```{r save-03-spatial-comparison, include=FALSE}
ggsave(here("figures", "03_qc_comparison_spatial.png"),
       sp_lenient + sp_standard + sp_strict,
       width = 18, height = 7, dpi = 300)
```

---

## 7. Key Observations

Numbers below are pulled directly from `qc_comparison` (section 5) rather
than hardcoded, so this section can't drift out of sync with the table above
on a re-run.

```{r key-obs-values, include=FALSE}
get_val <- function(cond, col) qc_comparison[qc_comparison$condition == cond, col]
```

- **Cells retained:** Lenient `r format(get_val("Lenient", "cells_retained"), big.mark = ",")` (`r get_val("Lenient", "pct_of_raw")`% of raw), Standard `r format(get_val("Standard", "cells_retained"), big.mark = ",")` (`r get_val("Standard", "pct_of_raw")`% of raw), Strict `r format(get_val("Strict", "cells_retained"), big.mark = ",")` (`r get_val("Strict", "pct_of_raw")`% of raw)
- **Cluster count:** Lenient `r get_val("Lenient", "n_clusters")`, Standard `r get_val("Standard", "n_clusters")`, Strict `r get_val("Strict", "n_clusters")`
- **Shared cluster IDs:** The union of cluster IDs across all conditions is used for a consistent
  color key. Cluster IDs are not biologically matched across conditions.
- **Interpretation:** More aggressive filtering trades cell yield for confidence per cell. It
  does not necessarily reduce the number of clusters recovered. Compare the cluster counts above
  directly: a stricter threshold may resolve as many (or more) distinct populations from fewer,
  cleaner cells if it is removing noise that was previously diluting cluster signal.
- **Recommendation:** Standard (nCount > 25, nFeature > 20) offers the best balance of
  cell yield and cluster resolution for Xenium 5K breast cancer data, based on the comparison
  above.

---

## 8. Save Objects

```{r save-objects, eval=FALSE}
saveRDS(obj_lenient,  here("data", "xenium_lenient.rds"))
saveRDS(obj_standard, here("data", "xenium_standard.rds"))
saveRDS(obj_strict,   here("data", "xenium_strict.rds"))
cat("All three objects saved to data/\n")
```

---

## 9. Figures Generated

| File | Description |
|------|-------------|
| `03_qc_comparison_summary.csv` | Summary table (4 conditions) |
| `03_qc_comparison_umaps.png` | 3-way UMAP comparison |
| `03_qc_comparison_spatial.png` | 3-way spatial comparison |
| `03_spatial_lenient.png` | Lenient spatial plot |
| `03_spatial_standard.png` | Standard spatial plot |
| `03_spatial_strict.png` | Strict spatial plot |

---

## 10. Session Info

```{r session-info}
sessionInfo()
```
