---
title: "04 - Spatial Visualization"
author: "Adrienne Yanez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

This notebook takes the standard filtered object (the recommended threshold
from `03_qc_comparison.Rmd`) and addresses the biological follow-up question:
where do these clusters sit in the tissue, and do they align with canonical
breast cancer cell-type markers (epithelial/tumor, T cell, B cell, myeloid,
endothelial, stromal)? Spatial coherence of a cluster, where cells of one type
form a plausible histological structures rather than being scattered
randomly, is an initial sanity check on the clustering, independent of marker
expression.

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_knit$set(root.dir = here::here())

library(Seurat)
library(ggplot2)
library(patchwork)
library(viridis)
library(dplyr)
library(here)

catalyst_cols <- c(
  "#7BAFDE",
  "#B2DF8A",
  "#BEAED4",
  "#8DD3C7",
  "#DC050C",
  "#FF7F00",
  "#882E72",
  "#33A02C",
  "#E7298A",
  "#E6AB02",
  "#1965B0",
  "#55A1B1",
  "#7570B3",
  "#FB8072",
  "#E78AC3",
  "#B17BA6",
  "#FDB462",
  "#A6761D",
  "#666666"
)

fix_clusters <- function(obj, col = "cluster_full") {
  vals <- as.character(obj[[col, drop = TRUE]])
  lvls <- as.character(sort(as.integer(unique(vals))))
  obj[[col]] <- factor(vals, levels = lvls)
  pal <- setNames(catalyst_cols[seq_along(lvls)], lvls)
  list(obj = obj, pal = pal, levels = lvls)
}
```

---

## Overview

Spatial visualization of the **standard-filtered object** (`nCount > 25,
nFeature > 20`) -the recommended QC threshold from notebook 03. Sections:

1. Full tissue cluster overlay
2. UMAP + spatial side by side
3. Canonical breast cancer marker genes, spatial
4. Marker genes, UMAP
5. Sub-region zoom
6. Multipanel cluster highlight (3 clusters at a time)
7. UMAP density plot
8. Cluster proportions stacked bar

---

## 1. Load Standard Object

```{r load-object, message=FALSE}
obj <- readRDS(here("data", "xenium_standard.rds"))
DefaultAssay(obj) <- "Xenium"

res <- fix_clusters(obj, col = "cluster_full")
obj <- res$obj
pal <- res$pal

cat("Cells:", ncol(obj), "\n")
cat("Cluster levels:", res$levels, "\n")
cat("Clusters with cells:", sum(table(obj$cluster_full) > 0), "\n")
scales::show_col(pal)
```

---

## 2. Full tissue clusters spatial plot

```{r spatial-clusters, fig.width=10, fig.height=8, message=FALSE, warning=FALSE}
DefaultAssay(obj) <- "Xenium"

ImageDimPlot(obj,
             group.by = "cluster_full",
             cols     = pal,
             size     = 0.5) +
  ggtitle("Spatial Clusters Standard QC Filtered\n(nCount > 25, nFeature > 20)") +
  theme(plot.title = element_text(size = 13))
```

```{r save-04-spatial-clusters, include=FALSE}
ggsave(here("figures", "04_spatial_clusters_standard.png"),
       ImageDimPlot(obj, group.by = "cluster_full", cols = pal, size = 0.5) +
         ggtitle("Spatial Clusters Standard QC Filtered"),
       width = 10, height = 8, dpi = 300)
```

---

## 3. UMAP and spatial 

```{r umap-spatial-combined, fig.width=16, fig.height=7, message=FALSE, warning=FALSE}
DefaultAssay(obj) <- "sketch"
p_umap <- DimPlot(obj, reduction = "umap", label = TRUE,
                  cols = pal, label.size = 4, pt.size = 0.4) +
  ggtitle("UMAP Standard Filtering") +
  theme_minimal() + NoLegend()

DefaultAssay(obj) <- "Xenium"
p_spatial <- ImageDimPlot(obj, group.by = "cluster_full",
                          cols = pal, size = 0.5) +
  ggtitle("Spatial Standard Filtering") + NoLegend()

p_umap + p_spatial
```

```{r save-04-umap-spatial, include=FALSE}
ggsave(here("figures", "04_umap_spatial_standard.png"),
       p_umap + p_spatial,
       width = 16, height = 7, dpi = 300)
```

---

## 4. Canonical breast cancer marker genes spatial

Zero expressing cells plotted as dark grey tissue background; expressing
cells overlaid in ascending order (high expressers on top). Coordinates rotated to match
tissue orientation (x = y, y = x).

```{r marker-genes-spatial, fig.width=9.5, fig.height=10, message=FALSE, warning=FALSE}
DefaultAssay(obj) <- "Xenium"

# Canonical lineage markers: EPCAM (epithelial/tumor), CD3E (T cell),
# MS4A1 (B cell), CD68 (myeloid/macrophage), PECAM1 (endothelial),
# ACTA2 (stromal/smooth muscle) -used to sanity-check that spatial
# clusters correspond to expected breast tissue cell types.
markers           <- c("EPCAM", "CD3E", "MS4A1", "CD68", "PECAM1", "ACTA2")
available_markers <- markers[markers %in% rownames(obj)]
cat("Available markers:", paste(available_markers, collapse = ", "), "\n")

plots <- lapply(available_markers, function(gene) {
  coords  <- GetTissueCoordinates(obj)
  expr    <- FetchData(obj, vars = gene)
  plot_df <- data.frame(x = coords$x, y = coords$y, expr = expr[[gene]])

  plot_zero    <- plot_df[plot_df$expr == 0, ]
  plot_express <- plot_df[plot_df$expr > 0, ]
  plot_express <- plot_express[order(plot_express$expr), ]
  q95 <- quantile(plot_express$expr, 0.95)
  plot_express$expr_capped <- pmin(plot_express$expr, q95)

  ggplot() +
    geom_point(data = plot_zero,
               aes(x = y, y = x),
               color = "#2d2d2d", size = 0.3, alpha = 0.5) +
    geom_point(data = plot_express,
               aes(x = y, y = x, color = expr_capped),
               size = 0.6) +
    scale_color_viridis_c(option = "magma", name = gene) +
    ggtitle(gene) +
    theme_void() +
    theme(plot.title      = element_text(hjust = 0.5, size = 12, face = "bold"),
          plot.background = element_rect(fill = "white", color = NA),
          legend.text     = element_text(size = 8),
          legend.title    = element_text(size = 9))
})

p_markers <- wrap_plots(plots, ncol = 2) &
  theme(plot.background = element_rect(fill = "white", color = NA))
p_markers
```

```{r save-04-markers-spatial, include=FALSE}
ggsave(here("figures", "04_spatial_markers.png"), p_markers,
       width = 9.5, height = 10, dpi = 300, bg = "white")
```

---

## 5. Marker Gene Expression UMAP

```{r marker-genes-umap, fig.width=9.5, fig.height=10, message=FALSE, warning=FALSE}
DefaultAssay(obj) <- "sketch"

FeaturePlot(obj,
            features   = available_markers,
            max.cutoff = "q95",
            pt.size    = 0.2) +
  plot_annotation(title = "Canonical Cell Type Markers UMAP")
```

```{r save-04-markers-umap, include=FALSE}
DefaultAssay(obj) <- "sketch"
ggsave(here("figures", "04_umap_markers.png"),
       FeaturePlot(obj, features = available_markers,
                   max.cutoff = "q95", pt.size = 0.2) +
         plot_annotation(title = "Canonical Cell Type Markers UMAP"),
       width = 9.5, height = 10, dpi = 300)
```


---

## 6. Multipanel Cluster Highlight (3 at a time)

Selected clusters colored, all others shown in dark grey for tissue context.

```{r multipanel-clusters, fig.width=14, fig.height=18, message=FALSE, warning=FALSE}
DefaultAssay(obj) <- "Xenium"
cluster_levels <- levels(obj$cluster_full)
cluster_groups <- split(cluster_levels,
                        ceiling(seq_along(cluster_levels) / 3))

panels <- lapply(cluster_groups, function(group_clusters) {
  group_pal        <- ifelse(names(pal) %in% group_clusters, pal, "#444444")
  names(group_pal) <- names(pal)

  ImageDimPlot(obj,
               group.by = "cluster_full",
               cols     = group_pal,
               size     = 0.5) +
    ggtitle(paste("Clusters:", paste(group_clusters, collapse = ", "))) +
    theme(plot.title = element_text(size = 11, hjust = 0.5)) +
    NoLegend()
})

wrap_plots(panels, ncol = 2) +
  plot_annotation(
    title    = "Standard Filtering",
    subtitle = "nCount > 25, nFeature > 20 | n=454,652 cells",
    theme    = theme(plot.title    = element_text(size = 14, face = "bold"),
                     plot.subtitle = element_text(size = 11))
  )
```

```{r save-04-multipanel, include=FALSE}
ggsave(here("figures", "04_spatial_clusters_multipanel.png"),
       wrap_plots(panels, ncol = 2) +
         plot_annotation(
           title    = "Standard Filtering",
           subtitle = "nCount > 25, nFeature > 20 | n=454,652 cells"
         ),
       width = 14, height = 18, dpi = 300)
```

---

## 7. UMAP Density Plot

```{r umap-density, fig.width=6, fig.height=5, message=FALSE, warning=FALSE}
DefaultAssay(obj) <- "sketch"
umap_coords <- as.data.frame(Embeddings(obj, "umap"))
colnames(umap_coords) <- c("UMAP_1", "UMAP_2")

umap.density.plot <- ggplot(umap_coords, aes(x = UMAP_1, y = UMAP_2)) +
  stat_density_2d_filled(contour_var = "ndensity", bins = 10,
                         adjust = 0.75, alpha = 0.9, show.legend = TRUE) +
  scale_fill_viridis(option = "A", discrete = TRUE, name = "Density") +
  labs(subtitle = "Cell Density (Standard Filtering)") +
  theme_minimal() +
  theme(plot.subtitle = element_text(hjust = 0.5))

umap.density.plot
```

```{r save-04-umap-density, include=FALSE}
ggsave(here("figures", "04_umap_density_standard.png"), umap.density.plot,
       width = 6, height = 5, dpi = 300)
```

---

## 8. Standalone Color Legend

```{r color-legend, fig.width=10, fig.height=2, message=FALSE, warning=FALSE}
legend_df <- data.frame(
  cluster = factor(names(pal), levels = names(pal)),
  x = 1, y = 1
)

ggplot(legend_df, aes(x = x, y = y, fill = cluster)) +
  geom_point(shape = 21, size = 5) +
  scale_fill_manual(values = pal, name = "Cluster",
                    guide = guide_legend(nrow = 2, reverse = FALSE,
                                        override.aes = list(size = 5))) +
  theme_void() +
  theme(legend.position = "bottom",
        legend.direction = "horizontal",
        legend.text  = element_text(size = 11),
        legend.title = element_text(size = 12, face = "bold"))
```

```{r save-04-legend, include=FALSE}
ggsave(here("figures", "04_cluster_color_legend.png"),
       last_plot(),
       width = 10, height = 2, dpi = 300, bg = "white")
```

---

## 9. Figures Generated

| File | Description | Used in |
|------|-------------|---------|
| `04_spatial_clusters_standard.png` | Full tissue, standard filtering | Slide 5 |
| `04_umap_spatial_standard.png` | UMAP + spatial, matched colors | Slide 5 |
| `04_spatial_markers.png` | 5 marker genes on tissue | Slide 5 |
| `04_umap_markers.png` | 5 marker genes on UMAP | Slide 5 |
| `04_spatial_clusters_multipanel.png` | 3-cluster highlight panels | Slide 5 |
| `04_umap_density_standard.png` | UMAP density overlay | Slide 5 (optional) |
| `04_cluster_color_legend.png` | Standalone cluster color key | Slides 5 and 6 |

---

## 10. Session Info

```{r session-info}
sessionInfo()
```
